Geometrical and physical interpretation of the action principle

### 2022-07-29 TeachingMaterial/ActionPrinciple/ActionPrincipleArticle.tex
ActionPrinciple: using square brackets for components

As we want to characterize the evolution of states over time, the appropriate setting is phase space extended with the time variable. That is, the space charted by position $q$, momentum $p$ and time $t$ as can be seen in Fig. \ref{extended_phase_space}. In the same way that we write $x^i = \{ x, y, z \}$ for the three dimension of space, we write
\begin{equation}\label{sdof_variables}
	\xi^a = \{ q, p, t\}
\end{equation}
for the three dimensions of the extended phase space.

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{the system undergoes deterministic and reversible evolution}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left\{ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right\} \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that tells how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} The time component of the displacement vector field is constrained as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

@@

Since the displacement field is divergence-free, it admits a vector potential. We have
\begin{equation}\label{sdof_displacement_potential}
\begin{aligned}
	\vec{\theta} &= \{\theta_q, \theta_p, \theta_t\} \\
	&= \theta_a e^a \\
	\vec{S} &= - \nabla \times \vec{\theta} \\
	&= - \epsilon^{abc} \partial_b \theta_c \, e_a. \\
\end{aligned}
\end{equation}
The minus sign is introduced to match conventions. Mathematically, this is analogous to what is done for a magnetic field or for an incompressible fluid.

Because the displacement field must satisfy \ref{sdof_time_constraint}, without loss of generality we can set
\begin{equation}\label{sdof_potential_expression}
\begin{aligned}
	\vec{\theta} &= \{p, 0, -H\} \\
	&= p e^q - H e^t.
\end{aligned}
\end{equation}
By applying definition \ref{sdof_displacement} and expanding \ref{sdof_displacement_potential} with \ref{sdof_potential_expression}, we have
\begin{equation}\label{sdof_Ham_eq}
	\left\{ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right\} = - \nabla \times \vec{\theta} = \left\{ \frac{\partial H}{\partial p},-\frac{\partial H}{\partial q}, 1 \right\},
\end{equation}
which yields Hamilton's equations. Note that the arguments works in reverse: any Hamiltonian system with one degree of freedom yields a divergence-free displacement field, and therefore satisfies \ref{assum_detrev}.

@@

To generalize to multiple degree of freedom we have to abandon the tools provided by vector calculus and embrace the ones provided by differential geometry. For $N$ independent degrees of freedom, we will have the $2N+1$ manifold charted by the variables
\begin{equation}\label{mdof_variables}
	\xi^a = \{ q^i, p_i, t\}.
\end{equation}
The displacement vector field will be
\begin{equation}\label{mdof_displacement}
	\begin{aligned}
		\vec{S} &= S^a e_a = \frac{d\xi^a}{dt} e_a =\frac{dq^i}{dt} e_{q^i} + \frac{dp_i}{dt} e_{p_i} + \frac{dt}{dt} e_t
	\end{aligned}
\end{equation}
which also satisfies \ref{sdof_time_constraint}.

@@

Similarly to the single d.o.f., without loss of generality we can set
\begin{equation}\label{mdof_potential_expression}
\begin{aligned}
	\theta &= \{p_i, 0, -H\} \\
	&= p_i e^{q^i} - H e^t.
\end{aligned}
\end{equation}
and calculate the components of the counting form
\begin{equation}\label{mdof_form_components}
	\omega_{ab} = (-\partial\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		-\partial_{q^j} H & -\partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

For the case of one degree of freedom we are going to use vector calculus. This is possible since the problem can be studied with three variables (position, momentum and time)
\begin{equation}
\tag*{(\ref{sdof_variables})}
	\xi^a = \{ q, p, t\}
\end{equation}
which is the only setting where vector calculus truly works. In this sense, nature gave us a hand as three dimensional spaces are much more intuitive for us, and we can therefore develop the ideas in a familiar context before generalizing.

@@

This requirement tells us two things. First, is that $X$ must be a differentiable manifold. To define densities over a set of variables and map them back and forward in time Jacobians must be defined, which means the derivative across state variables must be defined as well, in particular with respect with time, which plays the double role of a coordinate of $X$ and the evolution parameter. This means that the evolutions must be differentiable curves and therefore $\vec{S}$ must be defined which leads to
\begin{equation}
\tag*{(\ref{sdof_displacement})}
	\begin{aligned}
		\vec{S} &= \left\{ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right\} \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
The double role of time gives us
\begin{equation}
\tag*{(\ref{sdof_time_constraint})}
 	S^t=\frac{dt}{dt}=1.
 \end{equation}

@@

For (ii), since $\vec{S}$ is divergence-free, we can find a vector potential
\begin{equation}
\tag*{(\ref{sdof_displacement_potential})}
	\begin{aligned}
		\vec{\theta} &= \{\theta_q, \theta_p, \theta_t\} \\
		&= \theta_a e^a \\
		\vec{S} &= - \nabla \times \vec{\theta} \\
		&= - \epsilon^{abc} \partial_b \theta_c \, e_a. \\
	\end{aligned}
\end{equation}
Note that the vector potential is not uniquely identified, since $\nabla \times(\vec{\theta} + \nabla f) = \nabla \times \vec{\theta}$ and therefore the displacement field $\vec{S}$ remains unchanged. This is what is commonly referred to as gauge freedom in physics, and it is analogous to what happens for a magnetic field or for an incompressible fluid.

We use this gauge freedom to set the momentum component to zero. In fact, we can always choose $f(q,p,t)$ such that $\partial_p f = -\theta_p$ and redefine $\vec{\theta}$ to be $\vec{\theta} + \nabla f$. Therefore we can set
\begin{equation}
	\vec{\theta} = \{\theta_q, 0, \theta_t\}
\end{equation}
without loss of generality.

We now use the constraint \ref{sdof_time_constraint} on the time component. We have
\begin{align}
	S^t = \frac{dt}{dt} &= - \left(\frac{\partial}{\partial q}  \theta_p - \frac{\partial}{\partial p}  \theta_q\right) \\
	&= - \left(\frac{\partial}{\partial q}  0 - \frac{\partial}{\partial p}  \theta_q\right) \\
	& = \frac{\partial \theta_q}{\partial p} = 1
\end{align}
Integrating, we have $\theta_q = p + g(q,t)$ where $g(q,t)$ is an arbitrary function. This function can be set to zero without loss of generality as it can be removed with a gauge transformation where $f(q,t)$ does not depend on $p$, and therefore $\theta_p$ will remain unchanged. Therefore we have:
\begin{equation}
	\vec{\theta} = \{p, 0, \theta_t\}
\end{equation}
Lastly, we rename the last component as $\theta_t = -H$ which leads to 
\begin{equation}
\tag*{(\ref{sdof_potential_expression})}
	\begin{aligned}
		\vec{\theta} &= \{p, 0, -H\} \\
		&= p e^q - H e^t.
	\end{aligned}
\end{equation}

>>>>

As we want to characterize the evolution of states over time, the appropriate setting is phase space extended with the time variable. That is, the space charted by position $q$, momentum $p$ and time $t$ as can be seen in Fig. \ref{extended_phase_space}. In the same way that we write $x^i = [ x, y, z ]$ for the three dimension of space, we write
\begin{equation}\label{sdof_variables}
	\xi^a = [ q, p, t]
\end{equation}
for the three dimensions of the extended phase space.

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{the system undergoes deterministic and reversible evolution}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that tells how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} The time component of the displacement vector field is constrained as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

@@

Since the displacement field is divergence-free, it admits a vector potential. We have
\begin{equation}\label{sdof_displacement_potential}
\begin{aligned}
	\vec{\theta} &= [\theta_q, \theta_p, \theta_t] \\
	&= \theta_a e^a \\
	\vec{S} &= - \nabla \times \vec{\theta} \\
	&= - \epsilon^{abc} \partial_b \theta_c \, e_a. \\
\end{aligned}
\end{equation}
The minus sign is introduced to match conventions. Mathematically, this is analogous to what is done for a magnetic field or for an incompressible fluid.

Because the displacement field must satisfy \ref{sdof_time_constraint}, without loss of generality we can set
\begin{equation}\label{sdof_potential_expression}
\begin{aligned}
	\vec{\theta} &= [p, 0, -H] \\
	&= p e^q - H e^t.
\end{aligned}
\end{equation}
By applying definition \ref{sdof_displacement} and expanding \ref{sdof_displacement_potential} with \ref{sdof_potential_expression}, we have
\begin{equation}\label{sdof_Ham_eq}
	\left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] = - \nabla \times \vec{\theta} = \left[ \frac{\partial H}{\partial p},-\frac{\partial H}{\partial q}, 1 \right],
\end{equation}
which yields Hamilton's equations. Note that the arguments works in reverse: any Hamiltonian system with one degree of freedom yields a divergence-free displacement field, and therefore satisfies \ref{assum_detrev}.

@@

To generalize to multiple degree of freedom we have to abandon the tools provided by vector calculus and embrace the ones provided by differential geometry. For $N$ independent degrees of freedom, we will have the $2N+1$ manifold charted by the variables
\begin{equation}\label{mdof_variables}
	\xi^a = [ q^i, p_i, t].
\end{equation}
The displacement vector field will be
\begin{equation}\label{mdof_displacement}
	\begin{aligned}
		\vec{S} &= S^a e_a = \frac{d\xi^a}{dt} e_a =\frac{dq^i}{dt} e_{q^i} + \frac{dp_i}{dt} e_{p_i} + \frac{dt}{dt} e_t
	\end{aligned}
\end{equation}
which also satisfies \ref{sdof_time_constraint}.

@@

Similarly to the single d.o.f., without loss of generality we can set
\begin{equation}\label{mdof_potential_expression}
\begin{aligned}
	\theta &= [p_i, 0, -H] \\
	&= p_i e^{q^i} - H e^t.
\end{aligned}
\end{equation}
and calculate the components of the counting form
\begin{equation}\label{mdof_form_components}
	\omega_{ab} = (-\partial\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		-\partial_{q^j} H & -\partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

For the case of one degree of freedom we are going to use vector calculus. This is possible since the problem can be studied with three variables (position, momentum and time)
\begin{equation}
\tag*{(\ref{sdof_variables})}
	\xi^a = [ q, p, t]
\end{equation}
which is the only setting where vector calculus truly works. In this sense, nature gave us a hand as three dimensional spaces are much more intuitive for us, and we can therefore develop the ideas in a familiar context before generalizing.

@@

This requirement tells us two things. First, is that $X$ must be a differentiable manifold. To define densities over a set of variables and map them back and forward in time Jacobians must be defined, which means the derivative across state variables must be defined as well, in particular with respect with time, which plays the double role of a coordinate of $X$ and the evolution parameter. This means that the evolutions must be differentiable curves and therefore $\vec{S}$ must be defined which leads to
\begin{equation}
\tag*{(\ref{sdof_displacement})}
	\begin{aligned}
		\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
The double role of time gives us
\begin{equation}
\tag*{(\ref{sdof_time_constraint})}
 	S^t=\frac{dt}{dt}=1.
 \end{equation}

@@

For (ii), since $\vec{S}$ is divergence-free, we can find a vector potential
\begin{equation}
\tag*{(\ref{sdof_displacement_potential})}
	\begin{aligned}
		\vec{\theta} &= [\theta_q, \theta_p, \theta_t] \\
		&= \theta_a e^a \\
		\vec{S} &= - \nabla \times \vec{\theta} \\
		&= - \epsilon^{abc} \partial_b \theta_c \, e_a. \\
	\end{aligned}
\end{equation}
Note that the vector potential is not uniquely identified, since $\nabla \times(\vec{\theta} + \nabla f) = \nabla \times \vec{\theta}$ and therefore the displacement field $\vec{S}$ remains unchanged. This is what is commonly referred to as gauge freedom in physics, and it is analogous to what happens for a magnetic field or for an incompressible fluid.

We use this gauge freedom to set the momentum component to zero. In fact, we can always choose $f(q,p,t)$ such that $\partial_p f = -\theta_p$ and redefine $\vec{\theta}$ to be $\vec{\theta} + \nabla f$. Therefore we can set
\begin{equation}
	\vec{\theta} = [\theta_q, 0, \theta_t]
\end{equation}
without loss of generality.

We now use the constraint \ref{sdof_time_constraint} on the time component. We have
\begin{align}
	S^t = \frac{dt}{dt} &= - \left(\frac{\partial}{\partial q}  \theta_p - \frac{\partial}{\partial p}  \theta_q\right) \\
	&= - \left(\frac{\partial}{\partial q}  0 - \frac{\partial}{\partial p}  \theta_q\right) \\
	& = \frac{\partial \theta_q}{\partial p} = 1
\end{align}
Integrating, we have $\theta_q = p + g(q,t)$ where $g(q,t)$ is an arbitrary function. This function can be set to zero without loss of generality as it can be removed with a gauge transformation where $f(q,t)$ does not depend on $p$, and therefore $\theta_p$ will remain unchanged. Therefore we have:
\begin{equation}
	\vec{\theta} = [p, 0, \theta_t]
\end{equation}
Lastly, we rename the last component as $\theta_t = -H$ which leads to 
\begin{equation}
\tag*{(\ref{sdof_potential_expression})}
	\begin{aligned}
		\vec{\theta} &= [p, 0, -H] \\
		&= p e^q - H e^t.
	\end{aligned}
\end{equation}
